---
title: "Geospatial Distribution Dynamics With griddy"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Geospatial Distribution Dynamics With griddy}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, message=FALSE, warning=FALSE}
library(griddy)
library(dplyr)
library(ggplot2)
library(sf)
library(spData)
library(spdep)
library(tidyr)
```

`griddy` starts from a long spatial panel: one row per place and period. The
first question is not model fitting. It is whether places moved through a
distribution and whether that movement differs by spatial context.

```{r}
data(us_states, package = "spData")
data(us_states_df, package = "spData")

states <- us_states |>
  left_join(us_states_df, by = c("NAME" = "state")) |>
  filter(!NAME %in% c("Alaska", "Hawaii", "Puerto Rico")) |>
  arrange(NAME)

panel <- states |>
  select(NAME, REGION, median_income_10, median_income_15, geometry) |>
  pivot_longer(
    starts_with("median_income_"),
    names_to = "year",
    values_to = "median_income"
  ) |>
  mutate(year = if_else(year == "median_income_10", 2010L, 2015L))
```

Classify values using pooled quantiles so class meanings are stable across
periods.

```{r, fig.alt="Heatmap of classic Markov transition probabilities between income quintiles."}
classes <- classify_dynamics(panel, NAME, year, median_income, k = 5)
classic <- markov_dynamics(classes, NAME, year, class)

classic
plot_transition_matrix(classic)
```

Spatial Markov conditions transitions on the starting class of the spatial lag.

```{r, fig.alt="Faceted heatmaps of spatial Markov transition probabilities by spatial-lag quintile."}
nb <- spdep::poly2nb(states, queen = TRUE)
spatial <- spatial_markov(panel, NAME, year, median_income, nb = nb, k = 5)

spatial$transitions |>
  select(id, from_time, to_time, lag_class, transition, spatial_lag) |>
  head()

lag_intervals(spatial)

transition_matrix(spatial, "count", lag_class = "Q1")

plot_spatial_markov(spatial)
```

Rank mobility is intentionally simple in v0.1: it is map-ready descriptive
movement, not the full PySAL rank-method suite.

```{r, fig.alt="Map of endpoint rank mobility across states, with higher values indicating upward rank movement."}
mobility <- rank_mobility(panel, NAME, year, median_income)

mobility |>
  st_drop_geometry() |>
  arrange(desc(abs_rank_change)) |>
  select(NAME, start_rank, end_rank, rank_change) |>
  head()

plot_rank_mobility(mobility)
```
