---
title: "Geospatial Distribution Dynamics With griddy"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Geospatial Distribution Dynamics With griddy}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
grd_eval_docs <- identical(Sys.getenv("IN_PKGDOWN"), "true")
knitr::opts_chunk$set(eval = grd_eval_docs)
```

```{r, message=FALSE, warning=FALSE}
library(griddy)
library(dplyr)
library(ggplot2)
library(sf)
library(sfdep)
library(spData)
library(tidyr)
```

`griddy` works on long spatial panels: one row per place per period, keyed by
explicit `id`, `time`, and `value` columns. The bundled `usjoin` panel — 48
contiguous US states, per-capita personal income, 1929 to 2009 — mirrors the
canonical PySAL `giddy` example used in Rey (2001).

```{r}
data(usjoin)

usjoin
```

## Classification

Pooled quantile classification cuts the entire panel into five bins and reuses
the same breakpoints for every year. That keeps class meanings comparable
across periods, which matters because year-specific quantiles can hide absolute
movement.

```{r}
classes <- classify_dynamics(usjoin, name, year, income, k = 5)

class_intervals(classes)
```

## Classic Markov

```{r, fig.alt="Heatmap of classic Markov transition probabilities for US state per-capita income, 1929 to 2009."}
classic <- markov_dynamics(classes, name, year, class)

classic
plot_transition_matrix(classic)
```

The diagonal dominance of the matrix — most states stay in their starting
quintile from one year to the next — is the headline finding from this panel
and recovers the result Rey (2001) reports.

## Spatial Markov

Spatial Markov conditions transition probabilities on the starting class of
each unit's spatial lag. The expectation is that being surrounded by rich (or
poor) neighbors shifts the transition odds beyond what the unit's own class
would predict.

```{r}
states <- us_states |>
  st_drop_geometry() |>
  filter(NAME %in% usjoin$name) |>
  pull(NAME)

geom <- us_states |>
  filter(NAME %in% usjoin$name) |>
  arrange(NAME) |>
  mutate(
    nb = st_contiguity(geometry),
    wt = st_weights(nb)
  )

panel <- usjoin |>
  filter(name %in% states) |>
  arrange(name, year)
```

`geometry` is the preferred way to pass spatial weights: an `sf` tibble with
one row per unit and `nb` / `wt` list-columns produced by `sfdep`. Carrying
weights as columns on the geography keeps the spatial frame, the weights, and
their row-standardization choice in one tidy object.

```{r, fig.alt="Faceted heatmaps of spatial Markov transition probabilities for US state per-capita income, by spatial-lag quintile."}
spatial <- spatial_markov(panel, name, year, income, geometry = geom, k = 5)

lag_intervals(spatial)
transition_matrix(spatial, "probability", lag_class = "Q1")

plot_spatial_markov(spatial)
```

Reading the facets left to right: states whose neighbors sit in the lowest
income lag class (Q1) are much stickier at the bottom of the distribution than
the pooled transition matrix would suggest. The spatial context conditioning is
what spatial Markov is built to surface.

## Rank Mobility

Rank mobility is intentionally simple in v0.1: an endpoint comparison of unit
ranks between the first and last observed periods. It is descriptive and
map-ready, not the full PySAL rank-method suite.

```{r, fig.alt="Map of endpoint rank mobility for US states, 1929 to 2009, with positive values indicating upward rank movement."}
mobility_panel <- usjoin |>
  filter(name %in% states) |>
  left_join(geom |> select(NAME), by = c("name" = "NAME")) |>
  st_as_sf()

mobility <- rank_mobility(mobility_panel, name, year, income)

mobility |>
  st_drop_geometry() |>
  arrange(desc(abs_rank_change)) |>
  select(name, start_rank, end_rank, rank_change) |>
  head()

plot_rank_mobility(mobility)
```

## Output schemas

Each result object exposes a tidy `transitions` table and a long `matrix` table
keyed by class labels. Inspecting them directly is the simplest way to learn
the schema.

```{r}
classic$transitions |> head()
spatial$transitions |> select(id, from_time, to_time, lag_class, transition, spatial_lag) |> head()
```

For programmatic access, every griddy object also has `as.data.frame()` and the
S3 generics `transition_matrix()`, `class_intervals()`, and (for spatial
objects) `lag_intervals()`.
