---
title: "Performance Benchmarks"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Performance Benchmarks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Performance matters because spatial Markov workflows can touch tract panels
with thousands of units and many periods. This vignette records the benchmark
shape. It is evaluated for pkgdown and skipped during package checks.

```{r setup, include=FALSE}
grd_eval_docs <- identical(Sys.getenv("IN_PKGDOWN"), "true")
knitr::opts_chunk$set(eval = grd_eval_docs)
```

```{r, message=FALSE, warning=FALSE}
library(griddy)
library(microbenchmark)
library(sf)
library(spdep)
library(dplyr)
library(tidyr)
```

## Synthetic grid panel

A 60 by 60 queen-contiguity grid over ten time periods exercises the same
classification and spatial-lag code paths a tract-scale panel would.

```{r}
make_panel <- function(nx = 60, ny = 60, years = 2010:2019) {
  cells <- st_make_grid(st_bbox(c(xmin = 0, ymin = 0, xmax = nx, ymax = ny)), n = c(nx, ny))
  grid <- st_sf(id = seq_along(cells), geometry = cells)

  panel <- tidyr::crossing(id = grid$id, year = years) |>
    left_join(st_drop_geometry(grid), by = "id") |>
    mutate(value = id + as.integer(factor(year)) + rnorm(n())) |>
    left_join(select(grid, id, geometry), by = "id") |>
    st_as_sf()

  list(grid = grid, panel = panel)
}

fx <- make_panel()
listw <- nb2listw(cell2nb(60, 60, type = "queen"), style = "W")

microbenchmark(
  classify = classify_dynamics(fx$panel, id, year, value, k = 5),
  markov = {
    cls <- classify_dynamics(fx$panel, id, year, value, k = 5)
    markov_dynamics(cls, id, year, class)
  },
  spatial = spatial_markov(fx$panel, id, year, value, listw = listw, k = 5),
  times = 5
)
```

## Comparison against estdaR

The next chunk runs only when `estdaR` is installed. It calls
`griddy::spatial_markov()` and `estdaR::sp.mkv()` on the same wide panel,
matched class breaks, and identical weights. Any difference at this point
reflects implementation overhead, not differing inputs.

```{r, eval=grd_eval_docs && requireNamespace("estdaR", quietly = TRUE)}
library(estdaR)

wide_panel <- fx$panel |>
  st_drop_geometry() |>
  pivot_wider(id_cols = id, names_from = year, values_from = value) |>
  arrange(id)

wide_mat <- as.matrix(wide_panel |> select(-id))

microbenchmark(
  griddy = spatial_markov(fx$panel, id, year, value, listw = listw, k = 5),
  estdaR = estdaR::sp.mkv(wide_mat, listw, classes = 5, fixed = TRUE),
  times = 5
)
```

`griddy` carries the long-format reshape, classification, and label
preservation overhead that `estdaR::sp.mkv()` skips by working directly on a
unit-by-period matrix. The expected pattern is that `griddy` is slower at this
panel size; how much slower is the relevant question. If the gap exceeds an
order of magnitude on tract-scale panels it should be revisited.
