---
title: "Performance Benchmarks"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Performance Benchmarks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Performance matters because spatial Markov workflows can touch tract panels
with thousands of units and many periods. This vignette records the benchmark
shape. It is evaluated for pkgdown and skipped during package checks.

```{r setup, include=FALSE}
grd_eval_docs <- identical(Sys.getenv("IN_PKGDOWN"), "true")
knitr::opts_chunk$set(eval = grd_eval_docs)
```

```{r, message=FALSE, warning=FALSE}
library(griddy)
library(microbenchmark)
library(sf)
library(spdep)
library(dplyr)
library(tidyr)

make_panel <- function(nx = 60, ny = 60, years = 2010:2019) {
  cells <- st_make_grid(st_bbox(c(xmin = 0, ymin = 0, xmax = nx, ymax = ny)), n = c(nx, ny))
  grid <- st_sf(id = seq_along(cells), geometry = cells)

  panel <- tidyr::crossing(id = grid$id, year = years) |>
    left_join(st_drop_geometry(grid), by = "id") |>
    mutate(value = id + as.integer(factor(year)) + rnorm(n())) |>
    left_join(select(grid, id, geometry), by = "id") |>
    st_as_sf()

  list(grid = grid, panel = panel)
}

fx <- make_panel()
nb <- cell2nb(60, 60, type = "queen")

microbenchmark(
  classify = classify_dynamics(fx$panel, id, year, value, k = 5),
  markov = {
    cls <- classify_dynamics(fx$panel, id, year, value, k = 5)
    markov_dynamics(cls, id, year, class)
  },
  spatial = spatial_markov(fx$panel, id, year, value, nb = nb, k = 5),
  times = 5
)
```

Benchmark comparisons against `estdaR` and `spdyn` should use matched wide data,
matched class breaks, and identical weights. Otherwise the benchmark measures
classification defaults more than implementation performance.
