---
title: "Comparing Against estdaR and spdyn"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Comparing Against estdaR and spdyn}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

`estdaR` (Vallone et al.) and `spdyn` (Loaiza) are prior R implementations of
spatial distribution dynamics methods. Both predate `griddy` and remain useful
as methodological comparators. `griddy` does not compete on methodology —
it targets a different interface: long `sf` panels, tidy outputs, and
map-ready objects, with the unit identity, time index, class labels, transition
labels, and spatial-lag intervals attached to the result.

This vignette shows that `griddy::spatial_markov()` produces the same
transition counts as `estdaR::sp.mkv()` and `spdyn::spMarkov()` on a shared
panel, and walks through where the conventions intentionally differ.

```{r setup, include=FALSE}
grd_eval_docs <- identical(Sys.getenv("IN_PKGDOWN"), "true")
knitr::opts_chunk$set(eval = grd_eval_docs)
```

```{r, message=FALSE, warning=FALSE}
library(griddy)
library(dplyr)
library(sf)
library(spData)
library(spdep)
library(tidyr)
```

## Shared panel

The bundled `usjoin` panel mirrors PySAL's reference dataset, so prior R
implementations and `griddy` are working from the same series.

```{r}
data(usjoin)

geom <- us_states |>
  filter(NAME %in% usjoin$name) |>
  arrange(NAME)

panel <- usjoin |>
  filter(name %in% geom$NAME) |>
  arrange(name, year)

listw <- nb2listw(poly2nb(geom, queen = TRUE), style = "W")

spatial <- spatial_markov(panel, name, year, income, listw = listw, k = 5)

spatial$transitions |>
  select(id, from_time, to_time, lag_class, transition, spatial_lag) |>
  head()

lag_intervals(spatial)
transition_matrix(spatial, "count", lag_class = "Q1")
```

## Cross-package agreement

The next chunk runs only when `estdaR` and `spdyn` are both installed. The
package test suite carries static fixtures from both so ordinary CI can
validate the spatial Markov counts without depending on either package.

```{r, eval=grd_eval_docs && requireNamespace("estdaR", quietly = TRUE) && requireNamespace("spdyn", quietly = TRUE)}
library(estdaR)
library(spdyn)

wide <- panel |>
  pivot_wider(id_cols = name, names_from = year, values_from = income) |>
  arrange(name)

wide_mat <- as.matrix(wide |> select(-name))
year_cols <- as.character(unique(panel$year))

estdar_sp <- estdaR::sp.mkv(wide_mat, listw, classes = 5, fixed = TRUE)
spdyn_sp <- spdyn::spMarkov(
  wide,
  listw,
  stateVars = year_cols,
  n.states = 5,
  pool = TRUE,
  std = FALSE
)

griddy_counts <- aperm(
  array(
    spatial$matrix$n,
    dim = c(5, 5, 5),
    dimnames = list(spatial$states, spatial$states, spatial$lag_states)
  ),
  c(2, 1, 3)
)

all.equal(unname(griddy_counts), unname(estdar_sp$Transitions))
all.equal(unname(griddy_counts), unname(spdyn_sp$t))
```

When the two prior implementations and `griddy` agree to within numerical
tolerance, all three calls return `TRUE`. Disagreements with both at once are
treated as bugs in `griddy`. Disagreements with one but not the other are
documented under "Conventions worth knowing" below.

## Conventions worth knowing

Three places where a naive cross-comparison can produce false disagreement:

- **Classification breaks.** `griddy` uses pooled quantiles by default;
  `estdaR` accepts `fixed = TRUE`/`FALSE` to choose; `spdyn` exposes `pool`.
  Match those before comparing matrices.
- **Standardization.** `spdyn::spMarkov(..., std = TRUE)` standardizes the
  panel by year mean before classifying. `griddy` does not standardize by
  default. Set `std = FALSE` for direct comparison.
- **Output shape.** `estdaR` returns a 3D array indexed `[from, to, lag]`.
  `spdyn` returns an object with a `$t` field of the same shape. `griddy`
  returns a long `matrix` tibble keyed by `(lag_class, from_state, to_state)`,
  which is what makes downstream `dplyr` work natural.

The package test suite uses static fixtures from `estdaR::sp.mkv()` and
`spdyn::spMarkov()` so ordinary CI can validate the spatial Markov counts
without installing either package. Live cross-checks run conditionally via
`skip_if_not_installed()`.
