---
title: "Comparing Against estdaR and spdyn"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Comparing Against estdaR and spdyn}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

`estdaR` and `spdyn` are prior R implementations of spatial distribution
dynamics methods. They are useful as methodological comparators and oracle
checks. `griddy` targets a different interface: long `sf` panels, tidy outputs,
and map-ready objects.

The main difference is not that `griddy` computes an unrelated statistic. It is
that the unit identity, time index, class labels, transition labels, and spatial
lag intervals stay attached to the result.

```{r, message=FALSE, warning=FALSE}
library(griddy)
library(dplyr)
library(sf)
library(spData)
library(spdep)
library(tidyr)

data(us_states, package = "spData")
data(us_states_df, package = "spData")

states <- us_states |>
  left_join(us_states_df, by = c("NAME" = "state")) |>
  filter(!NAME %in% c("Alaska", "Hawaii", "Puerto Rico")) |>
  arrange(NAME)

panel <- states |>
  select(NAME, median_income_10, median_income_15, geometry) |>
  pivot_longer(starts_with("median_income_"), names_to = "year", values_to = "income") |>
  mutate(year = if_else(year == "median_income_10", 2010L, 2015L))

listw <- nb2listw(poly2nb(states, queen = TRUE), style = "W")

spatial <- spatial_markov(panel, NAME, year, income, listw = listw, k = 5)

spatial$transitions |>
  select(id, from_time, to_time, lag_class, transition, spatial_lag) |>
  head()

lag_intervals(spatial)
transition_matrix(spatial, "count", lag_class = "Q1")
```

The code below is intentionally optional because neither package is a hard
dependency.

```{r, eval=FALSE}
library(griddy)
library(estdaR)
library(spdyn)
library(spData)
library(spdep)
library(dplyr)
library(tidyr)

data(us_states, package = "spData")
data(us_states_df, package = "spData")

states <- us_states |>
  left_join(us_states_df, by = c("NAME" = "state")) |>
  filter(!NAME %in% c("Alaska", "Hawaii", "Puerto Rico")) |>
  arrange(NAME)

wide <- states |>
  st_drop_geometry() |>
  select(NAME, median_income_10, median_income_15)

panel <- states |>
  select(NAME, median_income_10, median_income_15, geometry) |>
  pivot_longer(starts_with("median_income_"), names_to = "year", values_to = "income") |>
  mutate(year = if_else(year == "median_income_10", 2010L, 2015L))

nb <- poly2nb(states, queen = TRUE)
listw <- nb2listw(nb, style = "W")

grd <- panel |>
  classify_dynamics(NAME, year, income, k = 5) |>
  markov_dynamics(NAME, year, class)

estdar <- estdaR::mkv(as.matrix(wide[, c("median_income_10", "median_income_15")]), classes = 5, fixed = TRUE)
spdyn <- spdyn::markov(wide, stateVars = c("median_income_10", "median_income_15"), n.states = 5, pool = TRUE, std = FALSE)

transition_matrix(grd, "count")
estdar$Transitions
spdyn$t
```

Expected differences should be documented. Classification conventions and
standardization defaults differ across packages, so exact equality requires
matching those choices first. The package test suite uses static fixtures from
`estdaR::sp.mkv()` and `spdyn::spMarkov()` so ordinary CI can validate the
spatial Markov counts without installing either package.
