---
title: "Did This State Move, Or Did Its Spatial Context Change?"
output: html_document
---

```{r setup, message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
library(sf)
library(spData)
library(spdep)
library(tibble)
library(tidyr)

theme_set(theme_minimal(base_size = 12))
```

This prototype is intentionally not package code. It is a single-file workflow test using existing R infrastructure plus a small set of helpers. The question is whether the applied path feels clearer than wiring together `spdep`, generic Markov tables, and map joins by hand.

The example uses bundled `spData` state geometries and 2010/2015 median income. Two years is too short for substantive convergence claims, but it is enough to test the workflow surface: classification, classic transitions, spatially conditioned transitions, and map-ready mobility output.

```{r data}
data(us_states, package = "spData")
data(us_states_df, package = "spData")

states <- us_states |>
  left_join(us_states_df, by = c("NAME" = "state")) |>
  filter(!NAME %in% c("Alaska", "Hawaii", "Puerto Rico")) |>
  arrange(NAME)

state_panel <- states |>
  select(NAME, REGION, median_income_10, median_income_15, geometry) |>
  pivot_longer(
    cols = starts_with("median_income_"),
    names_to = "year",
    values_to = "median_income"
  ) |>
  mutate(year = if_else(year == "median_income_10", 2010L, 2015L)) |>
  arrange(NAME, year)

nb <- poly2nb(states, queen = TRUE)
lw <- nb2listw(nb, style = "W", zero.policy = TRUE)
```

```{r helpers}
pooled_breaks <- function(x, k = 5) {
  stats::quantile(x, probs = seq(0, 1, length.out = k + 1), na.rm = TRUE, type = 7)
}

classify_values <- function(x, breaks, labels = NULL) {
  if (is.null(labels)) labels <- paste0("Q", seq_len(length(breaks) - 1))
  cut(
    x,
    breaks = unique(breaks),
    labels = labels,
    include.lowest = TRUE,
    ordered_result = TRUE
  )
}

classify_dynamics <- function(data, id, time, value, k = 5, labels = NULL) {
  id <- rlang::ensym(id)
  time <- rlang::ensym(time)
  value <- rlang::ensym(value)

  breaks <- pooled_breaks(dplyr::pull(data, !!value), k = k)

  data |>
    mutate(
      class = classify_values(!!value, breaks = breaks, labels = labels),
      .before = !!value
    ) |>
    arrange(!!id, !!time) |>
    structure(
      class = c("grd_classes", class(data)),
      breaks = breaks,
      id = rlang::as_string(id),
      time = rlang::as_string(time),
      value = rlang::as_string(value)
    )
}

markov_dynamics <- function(classes, id, time, state = class) {
  id <- rlang::ensym(id)
  time <- rlang::ensym(time)
  state <- rlang::ensym(state)

  transitions <- classes |>
    st_drop_geometry() |>
    arrange(!!id, !!time) |>
    group_by(!!id) |>
    mutate(to_state = lead(!!state), to_time = lead(!!time)) |>
    ungroup() |>
    filter(!is.na(to_state)) |>
    transmute(
      from_time = !!time,
      to_time,
      from_state = !!state,
      to_state
    )

  states <- levels(droplevels(classes |> st_drop_geometry() |> pull(!!state)))
  counts <- table(
    factor(transitions$from_state, levels = states),
    factor(transitions$to_state, levels = states)
  )

  probabilities <- prop.table(counts, margin = 1)
  probabilities[is.nan(probabilities)] <- NA_real_

  list(
    transitions = transitions,
    counts = counts,
    probabilities = probabilities
  ) |>
    structure(class = "grd_markov")
}

spatial_markov <- function(panel, geography, listw, id, time, value, k = 5) {
  id <- rlang::ensym(id)
  time <- rlang::ensym(time)
  value <- rlang::ensym(value)

  ids <- geography |> st_drop_geometry() |> pull(!!id)
  value_breaks <- pooled_breaks(panel |> st_drop_geometry() |> pull(!!value), k = k)

  lag_panel <- panel |>
    st_drop_geometry() |>
    select(!!id, !!time, !!value) |>
    arrange(!!time, match(!!id, ids)) |>
    group_by(!!time) |>
    mutate(
      spatial_lag = lag.listw(listw, !!value, zero.policy = TRUE),
      value_class = classify_values(!!value, breaks = value_breaks)
    ) |>
    ungroup()

  lag_breaks <- pooled_breaks(lag_panel$spatial_lag, k = k)

  transitions <- lag_panel |>
    mutate(lag_class = classify_values(spatial_lag, breaks = lag_breaks)) |>
    arrange(!!id, !!time) |>
    group_by(!!id) |>
    mutate(
      to_value_class = lead(value_class),
      to_time = lead(!!time)
    ) |>
    ungroup() |>
    filter(!is.na(to_value_class)) |>
    transmute(
      !!id,
      from_time = !!time,
      to_time,
      lag_class,
      from_state = value_class,
      to_state = to_value_class
    )

  state_levels <- paste0("Q", seq_len(k))
  lag_levels <- paste0("Q", seq_len(k))

  matrices <- transitions |>
    count(lag_class, from_state, to_state, name = "n") |>
    complete(
      lag_class = factor(lag_levels, levels = lag_levels, ordered = TRUE),
      from_state = factor(state_levels, levels = state_levels, ordered = TRUE),
      to_state = factor(state_levels, levels = state_levels, ordered = TRUE),
      fill = list(n = 0)
    ) |>
    group_by(lag_class, from_state) |>
    mutate(probability = if (sum(n) > 0) n / sum(n) else NA_real_) |>
    ungroup()

  list(
    transitions = transitions,
    matrices = matrices,
    value_breaks = value_breaks,
    lag_breaks = lag_breaks
  ) |>
    structure(class = "grd_spatial_markov")
}

rank_mobility <- function(panel, id, time, value, compare = c("endpoint", "adjacent")) {
  compare <- match.arg(compare)
  id <- rlang::ensym(id)
  time <- rlang::ensym(time)
  value <- rlang::ensym(value)

  ranked <- panel |>
    group_by(!!time) |>
    mutate(rank = min_rank(desc(!!value))) |>
    ungroup() |>
    arrange(!!id, !!time)

  if (compare == "endpoint") {
    endpoints <- ranked |>
      group_by(!!id) |>
      filter(!!time %in% range(!!time)) |>
      mutate(
        start_rank = first(rank),
        end_rank = last(rank),
        rank_change = start_rank - end_rank
      ) |>
      slice_tail(n = 1) |>
      ungroup()

    return(endpoints)
  }

  ranked |>
    group_by(!!id) |>
    mutate(rank_change = rank - lead(rank)) |>
    ungroup()
}

plot_transition_matrix <- function(markov) {
  as.data.frame(markov$probabilities) |>
    as_tibble(.name_repair = "minimal") |>
    rename(from_state = Var1, to_state = Var2, probability = Freq) |>
    ggplot(aes(to_state, from_state, fill = probability)) +
    geom_tile(color = "white", linewidth = 0.4) +
    geom_text(aes(label = scales::percent(probability, accuracy = 1)), na.rm = TRUE, size = 3) +
    scale_fill_viridis_c(labels = scales::percent, na.value = "grey90") +
    labs(x = "2015 class", y = "2010 class", fill = "Probability")
}

plot_spatial_markov <- function(spatial_markov) {
  spatial_markov$matrices |>
    ggplot(aes(to_state, from_state, fill = probability)) +
    geom_tile(color = "white", linewidth = 0.3) +
    facet_wrap(vars(lag_class), nrow = 1) +
    scale_fill_viridis_c(labels = scales::percent, na.value = "grey90") +
    labs(x = "2015 income class", y = "2010 income class", fill = "Probability")
}
```

## Classic Movement

```{r classify-and-markov}
classes <- classify_dynamics(
  state_panel,
  id = NAME,
  time = year,
  value = median_income,
  k = 5
)

classic <- markov_dynamics(classes, id = NAME, time = year)

classic$counts
classic$probabilities
```

```{r transition-heatmap, fig.width=6.5, fig.height=5}
plot_transition_matrix(classic)
```

## Spatial Context

```{r spatial-markov}
spatial <- spatial_markov(
  panel = state_panel,
  geography = states,
  listw = lw,
  id = NAME,
  time = year,
  value = median_income,
  k = 5
)

spatial$matrices |>
  filter(!is.na(probability), n > 0) |>
  arrange(lag_class, from_state, to_state)
```

```{r spatial-heatmap, fig.width=10, fig.height=3.8}
plot_spatial_markov(spatial)
```

The spatial Markov table conditions income-class movement on the starting class of neighboring states' income. In a longer panel, this is the workflow object that should carry the interpretation: not just "which states moved," but "whether movement differs when the surrounding context starts poorer or richer."

## Map-Ready Rank Mobility

```{r rank-mobility}
mobility <- rank_mobility(state_panel, id = NAME, time = year, value = median_income) |>
  select(NAME, year, median_income, rank, start_rank, end_rank, rank_change, geometry)

mobility |>
  st_drop_geometry() |>
  arrange(desc(abs(rank_change))) |>
  head(10)
```

```{r mobility-map, fig.width=8, fig.height=5}
ggplot(mobility) +
  geom_sf(aes(fill = rank_change), color = "white", linewidth = 0.2) +
  scale_fill_gradient2(
    low = "#b2182b",
    mid = "grey95",
    high = "#2166ac",
    midpoint = 0
  ) +
  labs(fill = "Rank gain")
```

## Prototype Notes

- The helper API is already more natural for long `sf` panel data than `estdaR`/`spdyn`-style wide matrices.
- `spdep` cleanly provides the spatial lag substrate. The package value has to come from the workflow object, validation, interpretation, and plots.
- A real article needs a longer panel. The next iteration should use county or tract ACS data cached as a fixture, not live API calls.
- The two-year state example is only a smoke test for the API shape.
