---
title: "Working With cancensus"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  eval = nzchar(Sys.getenv("CM_API_KEY")) ||
    nzchar(Sys.getenv("CANCENSUS_API_KEY"))
)
```

```{r, message=FALSE, warning=FALSE}
library(griddy)
library(dplyr)
```

This article is a website-only template for Canadian Census workflows: it
requires live CensusMapper access, so it is rendered on the pkgdown site when
the API key is set and skipped otherwise. It does not ship with the installed
package.

```{r, fig.alt=c("Heatmap of Canadian Census tract transition probabilities between income quintiles.", "Faceted heatmaps of Canadian Census tract spatial Markov transition probabilities by spatial-lag quintile.")}
library(cancensus)
library(sf)
library(spdep)

regions <- list(CSD = "5915022") # Vancouver CSD
vectors <- c(
  income_2016 = "v_CA16_2397",
  income_2021 = "v_CA21_906"
)

panel_2016 <- get_census(
  dataset = "CA16",
  regions = regions,
  vectors = vectors["income_2016"],
  level = "CT",
  geo_format = "sf"
) |>
  transmute(GeoUID, year = 2016L, median_income = income_2016, geometry)

panel_2021 <- get_census(
  dataset = "CA21",
  regions = regions,
  vectors = vectors["income_2021"],
  level = "CT",
  geo_format = "sf"
) |>
  transmute(GeoUID, year = 2021L, median_income = income_2021, geometry)

common_ids <- intersect(panel_2016$GeoUID, panel_2021$GeoUID)
panel <- bind_rows(panel_2016, panel_2021) |>
  filter(GeoUID %in% common_ids)
geography <- panel |> filter(year == 2021L) |> arrange(GeoUID)
listw <- nb2listw(poly2nb(geography, queen = TRUE), style = "W", zero.policy = TRUE)

classes <- classify_dynamics(panel, GeoUID, year, median_income, k = 5)
classic <- markov_dynamics(classes, GeoUID, year, class)
spatial <- spatial_markov(panel, GeoUID, year, median_income, listw = listw, k = 5)

plot_transition_matrix(classic)
plot_spatial_markov(spatial)
```

Canadian census geographies can change between cycles. A production workflow
should harmonize geography before treating units as a longitudinal panel.
