---
title: "Comparing Against estdaR and spdyn"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Comparing Against estdaR and spdyn}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

`estdaR` (Vallone et al.) and `spdyn` (Loaiza) are prior R implementations of
spatial distribution dynamics methods. Both predate `griddy` and remain useful
as methodological comparators. `griddy` does not compete on methodology —
it targets a different interface: long `sf` panels, tidy outputs, and
map-ready objects, with the unit identity, time index, class labels, transition
labels, and spatial-lag intervals attached to the result.

This vignette shows that `griddy::spatial_markov()` produces the same
transition counts as `estdaR::sp.mkv()` and `spdyn::spMarkov()` on a shared
panel, and walks through where the conventions intentionally differ.

```{r setup, include=FALSE}
grd_eval_docs <- identical(Sys.getenv("IN_PKGDOWN"), "true")
knitr::opts_chunk$set(eval = grd_eval_docs)
```

```{r, message=FALSE, warning=FALSE}
library(griddy)
library(dplyr)
library(sf)
library(sfdep)
library(spData)
library(spdep)
library(tidyr)
```

## Shared panel

The comparison uses the `spData` US states income panel because it is available
from CRAN and can be represented in the wide shape expected by both prior R
implementations. `estdaR` and `spdyn` consume `listw` objects from `spdep`;
`griddy` accepts the same object for compatibility, so all three methods use
identical row-standardized weights.

```{r}
data(us_states, package = "spData")
data(us_states_df, package = "spData")

states <- us_states |>
  left_join(us_states_df, by = c("NAME" = "state")) |>
  filter(!NAME %in% c("Alaska", "Hawaii", "Puerto Rico")) |>
  arrange(NAME)

panel <- states |>
  select(NAME, median_income_10, median_income_15, geometry) |>
  pivot_longer(
    starts_with("median_income_"),
    names_to = "year",
    values_to = "income"
  ) |>
  mutate(year = if_else(year == "median_income_10", 2010L, 2015L))

nb <- poly2nb(states, queen = TRUE)
listw <- nb2listw(nb, style = "W")

spatial <- spatial_markov(panel, NAME, year, income, listw = listw, k = 5)

lag_intervals(spatial)
transition_matrix(spatial, "count", lag_class = "Q1")
```

## Cross-package agreement

The next chunk uses live `estdaR` and `spdyn` calls when both packages are
installed. Otherwise, it compares against the committed oracle fixture produced
from those packages. This keeps the rendered site explicit about the agreement
without making the website build depend on packages that are not available from
CRAN.

```{r, eval=grd_eval_docs}
has_live_oracles <- requireNamespace("estdaR", quietly = TRUE) &&
  requireNamespace("spdyn", quietly = TRUE)

wide <- panel |>
  st_drop_geometry() |>
  pivot_wider(id_cols = NAME, names_from = year, values_from = income) |>
  arrange(NAME)

wide_mat <- as.matrix(wide |> select(-NAME))
year_cols <- as.character(unique(panel$year))

griddy_counts <- aperm(
  array(
    spatial$matrix$n,
    dim = c(5, 5, 5),
    dimnames = list(spatial$states, spatial$states, spatial$lag_states)
  ),
  c(2, 1, 3)
)

if (has_live_oracles) {
  estdar_sp <- estdaR::sp.mkv(wide_mat, listw, classes = 5, fixed = TRUE)
  spdyn_sp <- spdyn::spMarkov(
    wide,
    listw,
    stateVars = year_cols,
    n.states = 5,
    pool = TRUE,
    std = FALSE
  )

  estdaR_counts <- estdar_sp$Transitions
  spdyn_counts <- spdyn_sp$t
} else {
  fixture_path <- file.path("..", "..", "tests", "testthat", "fixtures", "r_prior_spatial_counts.rds")
  fixture <- readRDS(fixture_path)
  estdaR_counts <- fixture$estdaR_spatial_counts
  spdyn_counts <- fixture$spdyn_spatial_counts
}

all.equal(unname(griddy_counts), unname(estdaR_counts))
all.equal(unname(griddy_counts), unname(spdyn_counts))
```

When the two prior implementations and `griddy` agree to within numerical
tolerance, both calls return `TRUE`. Disagreements with both at once are treated
as bugs in `griddy`. Disagreements with one but not the other are documented
under "Conventions worth knowing" below.

## Conventions worth knowing

Three places where a naive cross-comparison can produce false disagreement:

- **Classification breaks.** `griddy` uses pooled quantiles by default;
  `estdaR` accepts `fixed = TRUE`/`FALSE` to choose; `spdyn` exposes `pool`.
  Match those before comparing matrices.
- **Standardization.** `spdyn::spMarkov(..., std = TRUE)` standardizes the
  panel by year mean before classifying. `griddy` does not standardize by
  default. Set `std = FALSE` for direct comparison.
- **Output shape.** `estdaR` returns a 3D array indexed `[from, to, lag]`.
  `spdyn` returns an object with a `$t` field of the same shape. `griddy`
  returns a long `matrix` tibble keyed by `(lag_class, from_state, to_state)`,
  which is what makes downstream `dplyr` work natural.

The package test suite uses static fixtures from `estdaR::sp.mkv()` and
`spdyn::spMarkov()` so ordinary CI can validate the spatial Markov counts
without installing either package. The live oracle script in `tools/oracle/`
can be run manually after installing both prior-art packages to refresh the
fixture.
